---
title: "millsratio: Mills Ratio Analysis in R"
subtitle: "Exploring tail thickness through Mills ratios and hazard functions"
format:
  html:
    toc: false
---

## Welcome

The `millsratio` package provides tools for analyzing and comparing Mills ratios across different distributions, revealing insights into tail thickness and extreme event probabilities.

## Key Features

:::: {.columns}
::: {.column width="50%"}
### Core Functions
- Calculate Mills ratios for multiple distributions
- Hazard function calculations
- Tail thickness analysis
- Interactive visualizations
:::

::: {.column width="50%"}
### Interactive Dashboard
- Explore Mills ratios interactively
- Compare distributions side-by-side
- Understand the hazard function connection
- Tidyverse code examples
:::
::::

## Quick Start

```{r}
#| eval: false
# Install from GitHub
remotes::install_github("JohnGavin/millsratio")

# Load the package
library(millsratio)
library(tidyverse)

# Calculate Mills ratios
x <- seq(0.1, 5, by = 0.1)
tibble(x = x) %>%
  mutate(
    mills_normal = mills_ratio_normal(x),
    mills_t30 = mills_ratio_t(x, df = 30),
    mills_exp = mills_ratio_exp(x)
  ) %>%
  pivot_longer(-x, names_to = "distribution", values_to = "mills_ratio")
```

## The t(30) Paradox

A key finding illustrated by this package:

> The t-distribution with 30 degrees of freedom appears nearly normal in the center but has substantially fatter tails, with the Mills ratio diverging increasingly at extreme values.

```{r}
#| eval: false
# Demonstrate the t(30) paradox
analyze_t30_paradox() %>%
  filter(x %in% c(1, 2, 3, 4)) %>%
  select(x, mills_ratio)
```

## Interactive Dashboard

Launch the interactive dashboard to explore Mills ratios and hazard functions:

```{r}
#| eval: false
# Launch the dashboard
launch_dashboard_simple()
```

Or try the web version (no R installation required):

<iframe src="shinylive/index.html" width="100%" height="600px"></iframe>

## Mathematical Foundation

The Mills ratio is defined as:

$$m(x) = \frac{1 - F(x)}{f(x)} = \frac{\bar{F}(x)}{f(x)}$$

where:
- $f(x)$ is the probability density function
- $F(x)$ is the cumulative distribution function
- $\bar{F}(x) = 1 - F(x)$ is the survival function

The hazard function is the reciprocal:

$$h(x) = \frac{1}{m(x)} = \frac{f(x)}{\bar{F}(x)}$$

## Learn More

- [Mathematical Theory](articles/theory.qmd)
- [Applications in Practice](articles/applications.qmd)
- [Hazard Function Connection](articles/hazard-connection.qmd)
- [Performance Benchmarks](articles/benchmarks.qmd)

## Citation

If you use this package, please cite:

```bibtex
@software{millsratio2026,
  author = {Gavin, John},
  title = {millsratio: Mills Ratio Analysis in R},
  year = {2026},
  url = {https://github.com/JohnGavin/millsratio}
}
```

Based on insights from [John D. Cook's blog post](https://www.johndcook.com/blog/2026/01/21/mills-ratio/) on Mills ratios and tail thickness.