---
title: "Interactive Dashboard"
subtitle: "Exploring Mills ratios through interactive visualization"
author: "John Gavin"
date: today
format:
  html:
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    toc: true
---

```{r setup}
#| message: false
#| warning: false
library(millsratio)
library(tidyverse)
library(plotly)
library(DT)
```

## Overview

The millsratio package includes two interactive dashboards for exploring Mills ratios and their relationship to distribution tail thickness:

1. **Simple Dashboard** (`launch_dashboard_simple()`) - Focused on core concepts
2. **Full Dashboard** (`launch_dashboard()`) - Comprehensive 12-page analysis

:::callout-note
## Running the Dashboard
The dashboards are **interactive Shiny applications** that run locally on your computer. They are not available on the website. To use them:

1. Install the millsratio package
2. Run `library(millsratio)` in R
3. Execute `launch_dashboard()` or `launch_dashboard_simple()`
4. Your browser will open with the interactive dashboard

The dashboards require a local R installation and cannot be embedded in static documentation.
:::

## Dashboard Features

### Simple Dashboard

The simplified dashboard focuses on essential functionality:

- **Analysis Page**: Compare Mills ratios across distributions
- **Theory Page**: Mathematical foundations and asymptotic behavior
- **Hazard Function Page**: Explore the h(x) = 1/m(x) relationship
- **R Code Page**: Ready-to-use tidyverse examples

Launch it with:

```{r}
launch_dashboard_simple()
```

### Full Dashboard (12 Pages)

The comprehensive dashboard is organized into three main sections:

#### Analysis Section
1. **Dashboard Overview**: Introduction and navigation guide
2. **Distribution Comparison**: Side-by-side analysis of multiple distributions
3. **The t(30) Paradox**: Deep dive into why t(30) appears normal but isn't
4. **Tail Thickness Analysis**: Quantitative measures of tail behavior

#### Theory Section
5. **Mathematical Foundation**: Definitions and derivations
6. **Asymptotic Behavior**: Long-run properties of Mills ratios
7. **Properties Comparison**: Table of distribution characteristics
8. **References**: Academic sources and further reading

#### Playground Section
9. **Interactive Explorer**: Real-time parameter adjustment
10. **Code Playground**: Write and test custom Mills ratio code
11. **Simulation Lab**: Monte Carlo verification tools
12. **Export Results**: Download analyses and visualizations

Launch it with:

```{r}
launch_dashboard()
```

## Key Interactive Features

### Real-time Visualization

Both dashboards provide instant feedback as you adjust parameters:

```{r}
#| fig-height: 6
# Example of what the dashboard visualizes
x <- seq(0.1, 5, by = 0.1)
data <- tibble(x = x) %>%
  mutate(
    Normal = mills_ratio_normal(x),
    `t(30)` = mills_ratio_t(x, df = 30),
    `t(3)` = mills_ratio_t(x, df = 3),
    Exponential = mills_ratio_exp(x)
  ) %>%
  pivot_longer(-x, names_to = "distribution", values_to = "mills_ratio")

p <- ggplot(data, aes(x, mills_ratio, color = distribution)) +
  geom_line(size = 1.2) +
  scale_y_log10() +
  labs(
    title = "Mills Ratio Comparison",
    subtitle = "Note the divergence in tail behavior",
    x = "x",
    y = "Mills Ratio m(x) (log scale)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplotly(p)
```

### Distribution Comparison Table

The dashboards generate comparison tables like this:

```{r}
# Compare Mills ratios at key points
x_points <- c(1, 2, 3, 4, 5)
comparison <- tibble(
  x = x_points,
  Normal = mills_ratio_normal(x_points),
  `t(30)` = mills_ratio_t(x_points, df = 30),
  `t(3)` = mills_ratio_t(x_points, df = 3),
  Exponential = mills_ratio_exp(x_points)
) %>%
  mutate(
    `t(30)/Normal` = `t(30)` / Normal,
    `t(3)/Normal` = `t(3)` / Normal
  )

comparison %>%
  DT::datatable(
    caption = "Mills Ratio Values and Ratios",
    options = list(pageLength = 5, dom = 't')
  ) %>%
  formatRound(2:7, digits = 4)
```

### The t(30) Paradox Visualization

A key feature demonstrates why t(30) is dangerous to treat as normal:

```{r}
#| fig-height: 6
# Analyze the paradox
paradox_data <- tibble(x = seq(0, 5, by = 0.1)) %>%
  mutate(
    normal = mills_ratio_normal(x),
    t30 = mills_ratio_t(x, df = 30),
    ratio = t30 / normal,
    percent_diff = (ratio - 1) * 100
  )

# Plot the divergence
p_paradox <- ggplot(paradox_data, aes(x, percent_diff)) +
  geom_line(color = "#e74c3c", size = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(
    title = "The t(30) Paradox: Percentage Difference from Normal",
    subtitle = "t(30) Mills ratio increasingly exceeds normal in the tails",
    x = "x",
    y = "% Difference from Normal"
  ) +
  theme_minimal() +
  annotate(
    "text", x = 4, y = 40,
    label = "At x=4: t(30) is 54% higher",
    color = "#e74c3c", fontface = "bold"
  )

ggplotly(p_paradox)
```

## Hazard Function Connection

The dashboard explores the reciprocal relationship between Mills ratio and hazard function:

```{r}
#| fig-height: 6
# Show both functions together
hazard_data <- tibble(x = seq(0.5, 5, by = 0.1)) %>%
  mutate(
    mills_normal = mills_ratio_normal(x),
    hazard_normal = 1 / mills_normal,
    mills_t30 = mills_ratio_t(x, df = 30),
    hazard_t30 = 1 / mills_t30
  ) %>%
  pivot_longer(
    -x,
    names_to = c("func_type", "distribution"),
    names_sep = "_",
    values_to = "value"
  )

p_hazard <- hazard_data %>%
  ggplot(aes(x, value, color = interaction(func_type, distribution))) +
  geom_line(size = 1) +
  scale_y_log10() +
  labs(
    title = "Mills Ratio m(x) and Hazard Function h(x) = 1/m(x)",
    x = "x",
    y = "Value (log scale)"
  ) +
  theme_minimal() +
  scale_color_manual(
    values = c(
      "mills.normal" = "#2c3e50",
      "hazard.normal" = "#e74c3c",
      "mills.t30" = "#3498db",
      "hazard.t30" = "#e67e22"
    ),
    labels = c(
      "mills.normal" = "Mills Normal",
      "hazard.normal" = "Hazard Normal",
      "mills.t30" = "Mills t(30)",
      "hazard.t30" = "Hazard t(30)"
    )
  ) +
  theme(legend.position = "bottom", legend.title = element_blank())

ggplotly(p_hazard)
```

## Code Examples in Dashboard

The dashboard includes ready-to-copy tidyverse code:

### Basic Mills Ratio Calculation

```r
library(millsratio)
library(tidyverse)

# Calculate and compare Mills ratios
df <- tibble(x = seq(0.5, 5, by = 0.5)) %>%
  mutate(
    normal = mills_ratio_normal(x),
    t30 = mills_ratio_t(x, df = 30),
    exponential = mills_ratio_exp(x),
    t30_normal_ratio = t30 / normal
  )
```

### Hazard Function Analysis

```r
# Hazard function from Mills ratio
df_hazard <- df %>%
  mutate(
    hazard_normal = 1 / normal,
    hazard_t30 = 1 / t30,
    hazard_exp = 1 / exponential
  )

# Check if hazard is increasing (IFR property)
df_hazard %>%
  arrange(x) %>%
  mutate(
    normal_increasing = hazard_normal > lag(hazard_normal),
    t30_increasing = hazard_t30 > lag(hazard_t30),
    exp_constant = abs(hazard_exp - lag(hazard_exp)) < 1e-10
  )
```

### Visualization Pipeline

```r
# Create comparison plot
df %>%
  select(x, normal, t30, exponential) %>%
  pivot_longer(-x, names_to = "distribution", values_to = "mills_ratio") %>%
  ggplot(aes(x, mills_ratio, color = distribution)) +
  geom_line(linewidth = 1) +
  scale_y_log10() +
  labs(
    title = "Mills Ratio Comparison",
    y = "Mills Ratio m(x) (log scale)"
  ) +
  theme_minimal()
```

## Dashboard Architecture

The dashboard is built with:

- **UI**: `bslib` for Bootstrap 5 theming
- **Server**: Reactive Shiny server with efficient caching
- **Plots**: `plotly` for interactive visualizations
- **Tables**: `DT` for sortable, searchable tables
- **Math**: MathJax for rendering mathematical notation

### File Structure

```
inst/shiny/
├── ui_simple.R         # Simplified UI (4 pages)
├── server_simple.R     # Simplified server logic
├── ui.R               # Full UI (12 pages)
├── server.R           # Full server logic
└── www/              # Static assets
    └── custom.css    # Custom styling
```

## Extending the Dashboard

To add new features to the dashboard:

1. **Add new distributions**: Implement in `R/mills_ratio.R`
2. **Add visualization**: Update `server.R` reactive functions
3. **Add theory content**: Edit `ui.R` nav_panel sections
4. **Add examples**: Update the Code Playground section

## Performance Considerations

The dashboards are optimized for performance:

- Reactive caching prevents redundant calculations
- Efficient vectorized operations throughout
- Lazy evaluation for expensive computations
- Debounced inputs to reduce update frequency

## Troubleshooting

Common issues and solutions:

1. **Dashboard won't launch**: Ensure all dependencies are installed
   ```r
   remotes::install_github("JohnGavin/millsratio")
   ```

2. **Plots not rendering**: Check that plotly is installed
   ```r
   install.packages("plotly")
   ```

3. **Math notation not displaying**: MathJax requires internet connection

4. **Slow performance**: Reduce the number of points in x-range

## Next Steps

After exploring the dashboard:

1. Read the [Mathematical Theory](theory.qmd) article
2. Review [Practical Applications](applications.qmd)
3. Explore the [Hazard Function Connection](hazard-connection.qmd)
4. Run the [Performance Benchmarks](benchmarks.qmd)
