---
title: "Mills Ratio and Hazard Function Connection"
subtitle: "Understanding the reciprocal relationship"
author: "John Gavin"
date: today
format:
  html:
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    output-fold: true
    output-summary: "Show output"
    toc: true
---

```{r setup}
#| message: false
#| warning: false
library(millsratio)
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
```

## Mathematical Relationship

The Mills ratio and hazard function are reciprocals of each other:

$$h(x) = \frac{f(x)}{1 - F(x)} = \frac{1}{m(x)}$$

This fundamental relationship connects two important concepts:

- **Mills ratio** $m(x)$: Used in statistics and economics to measure tail thickness
- **Hazard function** $h(x)$: Used in survival analysis and reliability engineering

## Numerical Verification: h(x) = 1/m(x) {#numerical-verification}

We verify the reciprocal identity numerically across all distributions.

```{r verification-table}
# Load precomputed verification table
verification <- readRDS(
  system.file("extdata", "hazard_verification_table.rds", package = "millsratio")
)

verification %>%
  select(distribution, x, mills_ratio, hazard, product, abs_error) %>%
  knitr::kable(
    digits = 12,
    caption = "Verification: m(x) * h(x) = 1 for all distributions and x values"
  )
```

The `product` column shows that $m(x) \times h(x) = 1$ to machine precision for every distribution and every $x$ value, confirming the reciprocal relationship.

## Side-by-Side Comparison {#side-by-side}

```{r side-by-side-plot}
#| fig-height: 6
# Load precomputed side-by-side data
side_by_side <- readRDS(
  system.file("extdata", "hazard_mills_side_by_side.rds", package = "millsratio")
)

plot_data <- side_by_side %>%
  pivot_longer(
    cols = c(mills_ratio, hazard),
    names_to = "function_type",
    values_to = "value"
  ) %>%
  mutate(
    function_type = ifelse(
      function_type == "mills_ratio",
      "Mills Ratio m(x)",
      "Hazard h(x)"
    )
  )

p <- ggplot(plot_data, aes(x, value, color = distribution)) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~function_type, scales = "free_y") +
  scale_y_log10(labels = scales::label_number()) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Mills Ratio vs Hazard Function",
    subtitle = "h(x) = 1/m(x) for all distributions",
    x = "x",
    y = "Value (log scale)",
    color = "Distribution"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  )

ggplotly(p)
```

Note how the patterns are mirror images on a log scale: where the Mills ratio is large (thick tails), the hazard is small, and vice versa.

## Compare Across Distributions {#compare-distributions}

```{r compare-distributions}
# Use precomputed side-by-side data for distribution comparison
p_compare <- side_by_side %>%
  ggplot(aes(x, hazard, color = distribution)) +
  geom_line(linewidth = 1) +
  scale_y_log10() +
  facet_wrap(~distribution, scales = "free_y") +
  labs(
    title = "Hazard Functions Across Distributions",
    x = "x",
    y = "Hazard h(x) (log scale)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggplotly(p_compare)
```

## Hazard Function Properties {#hazard-properties}

```{r hazard-properties}
# Load precomputed hazard properties
hazard_props <- readRDS(
  system.file("extdata", "hazard_properties.rds", package = "millsratio")
)

hazard_props %>%
  knitr::kable(
    caption = paste(
      "Hazard Function Properties by Distribution",
      "(IFR = Increasing Failure Rate, DFR = Decreasing, CFR = Constant)"
    )
  )
```

## Hazard Crossover: t(30) Paradox {#hazard-crossover}

The t(30) distribution is often treated as "approximately normal", but their hazard functions tell a different story. Despite nearly identical PDFs and CDFs, the hazard functions diverge in the tails.

```{r crossover-plot}
#| fig-height: 6
# Load precomputed crossover data
crossover <- readRDS(
  system.file("extdata", "hazard_crossover_data.rds", package = "millsratio")
)

# Find approximate crossover point
crossover_x <- crossover %>%
  mutate(diff_sign = sign(h_t30 - h_normal)) %>%
  mutate(sign_change = diff_sign != lag(diff_sign)) %>%
  filter(sign_change) %>%
  pull(x) %>%
  first()

p_cross <- ggplot(crossover, aes(x = x)) +
  geom_line(aes(y = h_normal, color = "Normal"), linewidth = 1.2) +
  geom_line(aes(y = h_t30, color = "t(30)"), linewidth = 1.2) +
  scale_color_manual(
    values = c("Normal" = "#1f77b4", "t(30)" = "#ff7f0e"),
    name = "Distribution"
  ) +
  scale_y_log10() +
  theme_minimal(base_size = 12) +
  labs(
    title = "Hazard Function Crossover: Normal vs t(30)",
    subtitle = sprintf("Crossover near x = %.1f", crossover_x),
    x = "x",
    y = "Hazard h(x) (log scale)"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  )

if (!is.na(crossover_x)) {
  p_cross <- p_cross +
    geom_vline(xintercept = crossover_x, linetype = "dashed", alpha = 0.5) +
    annotate("text", x = crossover_x + 0.3, y = 1, label = "Crossover",
             size = 3, alpha = 0.7)
}

ggplotly(p_cross)
```

```{r crossover-ratio}
#| fig-height: 4
# Ratio of hazard functions
p_ratio <- ggplot(crossover, aes(x = x, y = ratio)) +
  geom_line(linewidth = 1.2, color = "#2ca02c") +
  geom_hline(yintercept = 1, linetype = "dashed", alpha = 0.5) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Hazard Ratio: h_t30(x) / h_normal(x)",
    subtitle = "Values > 1 mean t(30) has higher hazard than normal",
    x = "x",
    y = "Hazard ratio"
  ) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggplotly(p_ratio)
```

For small $x$, the t(30) hazard exceeds the normal hazard (ratio > 1). In the tails, the normal hazard grows without bound while the t(30) hazard eventually decreases---a direct consequence of the heavier tails of the t-distribution.

## Applications {#applications}

### Survival Analysis Example

```{r survival-example}
# Compute survival curves from hazard functions
x_grid <- seq(0.1, 10, by = 0.1)
survival_data <- tibble(time = x_grid) %>%
  mutate(
    constant_hazard = hazard_function(time, "exponential", rate = 0.5),
    increasing_hazard = hazard_function(time, "normal", mean = 5, sd = 2),
    survival_constant = exp(-cumsum(constant_hazard * 0.1)),
    survival_increasing = exp(-cumsum(increasing_hazard * 0.1))
  )

p_survival <- survival_data %>%
  select(time, starts_with("survival")) %>%
  pivot_longer(-time, names_to = "type", values_to = "probability") %>%
  mutate(type = gsub("survival_", "", type)) %>%
  ggplot(aes(time, probability, color = type)) +
  geom_line(linewidth = 1.2) +
  labs(
    title = "Survival Curves from Different Hazard Functions",
    x = "Time",
    y = "Survival Probability",
    color = "Hazard Type"
  ) +
  theme_minimal()

ggplotly(p_survival)
```

## Key Insights {#key-insights}

1. **Reciprocal Relationship**: The hazard function is always the reciprocal of the Mills ratio: $h(x) = 1/m(x)$

2. **Distribution Classification**:
   - **Normal**: Increasing hazard (aging effect)
   - **Exponential**: Constant hazard (memoryless)
   - **t-distribution**: Eventually decreasing hazard (heavy tails)

3. **Practical Applications**:
   - Use Mills ratio for tail probability calculations
   - Use hazard function for time-to-event modeling
   - Both provide insights into extreme event behavior

## Code Examples {#code-examples}

### Pipeline for Analysis

```{r pipeline-example}
# Complete analysis pipeline using package functions
results <- tibble(
  distribution = c("Normal", "t(30)", "Exponential")
) %>%
  mutate(
    props = lapply(
      c("normal", "t", "exponential"),
      function(d) {
        if (d == "t") hazard_properties(d, df = 30)
        else hazard_properties(d)
      }
    ),
    behavior = sapply(props, `[[`, "behavior"),
    asymptotic = sapply(props, `[[`, "asymptotic_behavior"),
    interpretation = sapply(props, `[[`, "interpretation")
  ) %>%
  select(-props)

results %>%
  knitr::kable(caption = "Distribution Comparison Summary")
```

## Conclusion {#conclusion}

The Mills ratio and hazard function provide complementary views of the same underlying phenomenon. Understanding their relationship enables:

- Better interpretation of tail behavior
- Appropriate choice of distribution for modeling
- Cross-disciplinary insights between statistics and survival analysis

The `millsratio` package makes it easy to explore these relationships using tidyverse-style workflows.
