---
title: "Mathematical Theory of Mills Ratios"
subtitle: "Deep dive into the mathematics of tail thickness"
author: "John Gavin"
date: today
format:
  html:
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    toc: true
bibliography: ../references.bib
---

```{r setup}
#| message: false
#| warning: false
library(millsratio)
library(tidyverse)
library(plotly)
```

## Introduction

The Mills ratio, first tabulated by @mills1926, provides a fundamental measure of tail thickness in probability distributions. This article explores the mathematical foundations and theoretical properties of Mills ratios.

## Definition

The Mills ratio is defined as:

$$m(x) = \frac{\bar{F}(x)}{f(x)} = \frac{1 - F(x)}{f(x)}$$

where:
- $f(x)$ is the probability density function
- $F(x)$ is the cumulative distribution function
- $\bar{F}(x) = 1 - F(x)$ is the survival function

## Asymptotic Behavior

### Normal Distribution

For the standard normal distribution, as $x \to \infty$:

$$m(x) \sim \frac{1}{x} - \frac{1}{x^3} + \frac{3}{x^5} - \cdots$$

The leading term $1/x$ dominates, indicating thin tails.

```{r normal-asymptotic}
x <- seq(2, 10, by = 0.5)
mills_exact <- mills_ratio_normal(x)
mills_approx <- 1/x

data.frame(
  x = x,
  exact = mills_exact,
  approximation = mills_approx,
  relative_error = abs(mills_exact - mills_approx) / mills_exact
)
```

### Student's t-Distribution

For the t-distribution with $\nu$ degrees of freedom, as $x \to \infty$:

$$m(x) \sim \frac{x}{\nu}$$

This linear growth indicates fat tails.

### Exponential Distribution

For the exponential distribution with rate $\lambda$:

$$m(x) = \frac{1}{\lambda}$$

The constant Mills ratio characterizes the exponential tail decay.

## Inequalities and Bounds

### Sampford's Inequality

@sampford1953 established bounds for the normal Mills ratio:

$$\frac{1}{x + 1/x} < m(x) < \frac{1}{x}$$

These bounds become tighter as $x$ increases.

```{r bounds-visualization}
x <- seq(1, 5, by = 0.1)
mills <- mills_ratio_normal(x)
lower_bound <- 1/(x + 1/x)
upper_bound <- 1/x

plot_data <- data.frame(
  x = x,
  `Mills Ratio` = mills,
  `Lower Bound` = lower_bound,
  `Upper Bound` = upper_bound
) %>%
  pivot_longer(-x, names_to = "Type", values_to = "Value")

ggplot(plot_data, aes(x, Value, color = Type)) +
  geom_line(size = 1) +
  labs(
    title = "Mills Ratio Bounds for Normal Distribution",
    subtitle = "Sampford's inequality provides tight bounds",
    y = "Value"
  ) +
  theme_minimal()
```

## Relationship to Other Functions

### Hazard Function

The hazard function (failure rate) is the reciprocal of the Mills ratio:

$$h(x) = \frac{1}{m(x)} = \frac{f(x)}{\bar{F}(x)}$$

### Error Function

For the normal distribution:

$$m(x) = \frac{\sqrt{\pi/2} \cdot \text{erfc}(x/\sqrt{2})}{\exp(-x^2/2)}$$

where $\text{erfc}$ is the complementary error function.

## Tail Classification

Mills ratio behavior classifies tail thickness:

| Behavior | Tail Type | Example |
|----------|-----------|---------|
| Decreasing ($m'(x) < 0$) | Thin | Normal |
| Constant ($m'(x) = 0$) | Exponential | Exponential |
| Increasing ($m'(x) > 0$) | Fat | Student's t |

## Applications in Statistics

### Truncated Distributions

The mean of a truncated normal distribution above point $a$ is:

$$E[X | X > a] = \mu + \sigma \cdot \frac{1}{m((a-\mu)/\sigma)}$$

### Extreme Value Theory

Mills ratios appear in the study of:
- Order statistics
- Record values
- Peaks over threshold models

## Computational Considerations

### Numerical Stability

For large $x$, direct computation can be unstable. Use:

```{r numerical-tips}
# Unstable for large x
unstable_mills <- function(x) {
  pnorm(x, lower.tail = FALSE) / dnorm(x)
}

# Stable using log scale
stable_mills <- function(x) {
  exp(pnorm(x, lower.tail = FALSE, log.p = TRUE) - dnorm(x, log = TRUE))
}

# Compare at x = 10
x_test <- 10
c(unstable = unstable_mills(x_test),
  stable = stable_mills(x_test),
  package = mills_ratio_normal(x_test))
```

## References
